\scshape Advanced game theory

### 2025-12-07 ch_mcs.tex
Add files via upload

	The classical way of answering this question is to impose auxiliary differentiability and quasi-concavity assumptions, then use the first-order condition. In particular, assume additionally that $f$ is differentiable and strictly concave, and that it satisfies the Inada conditions $\lim_{a \downarrow 0} f'(a) = \infty$ and $\lim_{a \uparrow \infty} f'(a) = 0$. Then for each parameter $\theta \in \Theta$, there is a unique optimal choice $a^\star(\theta) \in \R_+$, which is interior (i.e. belongs to $(0,\infty)$) and satisfies the first-order condition
	%
	\begin{equation*}
		pzf'\left(a^\star(p,z,-r)\right) = r .
	\end{equation*}
	%
	Since $f'$ is strictly decreasing (as $f$ is strictly concave), it follows that $a^\star$ is increasing in each of its three arguments.

	This is all well and good, but what if $f$ fails to be differentiable or concave? The answer, as we shall see below, is that it makes no difference: all that need be assumed about the production function $f$ is that it is monotone.
	%
\end{example}

\begin{example}
	%
	\label{example:monopoly}
	%
	Consider a monopolist with constant marginal cost $c \in \R_+$ facing a decreasing and differentiable demand curve $D(\cdot,\eta) : \R_{++} \to \R_{++}$, where $\eta \in \R_+$ is a parameter that shifts the elasticity of demand: if $\eta \leq \eta'$ then the demand curve $D(\cdot,\eta)$ is uniformly less elastic than $D(\cdot,\eta')$, i.e.
	%
	\begin{equation*}
		- \frac{ D_1(a,\eta) a }{ D(a,\eta) }
		\leq
		- \frac{ D_1(a,\eta') a }{ D(a,\eta') }
		\quad \text{for every price $a \in \R_{++}$,}
	\end{equation*}
	%
	where `$D_1$' denotes the partial derivative of $D$ with respect to its first argument (namely, price). A simple parametric example is the iso-elastic demand curve $D(a,\eta) = K a^{-\eta}$ for all $a \in \R_{++}$, where $K>0$.
	The monopolist's problem is to choose a price $a \in A \coloneqq \R_{++}$ to maximise profit
	%
	\begin{equation*}
		u(a,c,-\eta) \coloneqq (a-c) \cdot D(a,\eta) .
	\end{equation*}

@@

@@

	\item The action set $A$ can be multi-dimensional: $A \subseteq \R^L$ for $L \in \N$. In this case, an additional assumption is required, which says that the different dimensions of the action $a \in A$ are complementary with each other (not only with the parameter $\theta$). One such condition (which is stronger than necessary, but easy to interpret) is \emph{supermodularity} of $u(\cdot,\theta)$, which assuming twice continuous differentiability requires precisely that for each $\theta \in \Theta$,
	%
	\begin{multline*}
		u_{\ell m}(a_1,\dots,a_\ell,\dots,a_m,\dots,a_L,\theta) \geq 0
		\\
		\text{for all $\ell \neq m$ in $\{1,\dots,L\}$ and all $(a_1,\dots,a_\ell,\dots,a_m,\dots,a_L) \in A$.}
	\end{multline*}
	%
	My own experience is that such supermodularity-type assumptions very rarely hold in economic applications, making this extension to multi-dimensional actions less useful than it sounds.

@@

	Recall the rudiments of dynamic programming \parencite[e.g.][]{StokeyLucasPrescott1989}. The Bellman equation is the following functional equation in an unknown function $f : \R_+ \to \R_+$:
	%
	\begin{equation*}
		f(w) =
		\max_{s \in [0,w]}
		\left[ U(w-s) + \delta \E\bigl( f\left( (1+R_1) s + Y_1 \right) \bigr) \right]
		\quad \text{for every $w \in \R_+$.}
	\end{equation*}
	%
	Under standard assumptions on $U$ and the distributions of $Y_1$ and $R_1$,%
		\footnote{For example, assume that $(Y_t,R_t)$ take values in compact set and that $U$ is bounded and continuous, and appeal to the results in \textcite[][chapter~9]{StokeyLucasPrescott1989}.}
	the value function $V : \R_+ \to \R_+$ solves the Bellman equation, and is the only solution that is bounded; furthermore, optimal saving policies $\sigma : \R_+ \to \R_+$ are precisely those that satisfy
	%
	\begin{equation*}
		\sigma(w) \in \argmax_{s \in [0,w]}
		\left[ U(w-s) + \delta \E\bigl( V\left( (1+R_1) s + Y_1 \right) \bigr) \right] 
		\quad \text{for every $w \in \R_+$.}
	\end{equation*}
	%
	Prove the following.

@@

\begin{exercise}[based on \textcite{DekelQuahSinander}]
	%
	\label{exercise:adj}
	%
	Consider a model of costly adjustment. An agent chooses $a \in A$, and her gross payoff is $u(a,\theta)$, where $\theta \in \Theta \subseteq \R^K$ is a parameter and $u$ has single-crossing differences. Initially, the parameter was $\theta_0 \in \Theta$, and the agent chose $a_0 \in \argmax_{a \in A} u(a,\theta_0)$. The parameter now increases to $\theta_1 \geq \theta_0$, and the agent adjusts her action in response. Adjustment is costly: in particular, adjusting by $\Delta = a_1 - a_0$ costs $c(\Delta) \geq 0$.  The agent chooses a new action to maximise her net payoff: $a_1 \in M_1 \coloneqq \argmax_{a \in A} [u(a,\theta_1)-c(a-a_0)]$. Assume that $M_1$ is non-empty.

>>>>

	The classical way of answering this question is to impose auxiliary differentiability and quasi-concavity assumptions, then use the first-order condition. In particular, assume additionally that $f$ is continuously differentiable and strictly concave, and that it satisfies the Inada conditions $\lim_{a \downarrow 0} f'(a) = \infty$ and $\lim_{a \uparrow \infty} f'(a) = 0$. Then for each parameter $\theta \in \Theta$, there is a unique optimal choice $a^\star(\theta) \in \R_+$, which is interior (i.e. belongs to $(0,\infty)$) and satisfies the first-order condition
	%
	\begin{equation*}
		pzf'\left(a^\star(p,z,-r)\right) = r .
	\end{equation*}
	%
	Since $f'$ is strictly decreasing (as $f$ is strictly concave), it follows that $a^\star$ is increasing in each of its three arguments.

	This is all well and good, but what if $f$ fails to be differentiable or concave? The answer, as we shall see below, is that it makes no difference: all that need be assumed about the production function $f$ is that it is increasing.
	%
\end{example}

\begin{example}
	%
	\label{example:monopoly}
	%
	Consider a monopolist with constant marginal cost $c \in \R_+$ facing a decreasing and differentiable demand curve $D(\cdot,\eta) : \R_{++} \to \R_{++}$, where $\eta \in \R_+$ is a parameter that shifts the elasticity of demand: if $\eta \leq \eta'$ then the demand curve $D(\cdot,\eta)$ is uniformly less elastic than $D(\cdot,\eta')$, i.e.
	%
	\begin{equation*}
		- \frac{ D_1(a,\eta) a }{ D(a,\eta) }
		\leq
		- \frac{ D_1(a,\eta') a }{ D(a,\eta') }
		\quad \text{for every price $a \in \R_{++}$,}
	\end{equation*}
	%
	where `$D_1$' denotes the partial derivative of $D$ with respect to its first argument (namely, price). A simple parametric example is the iso-elastic demand curve $D(a,\eta) = K a^{-\eta}$ for all $a \in \R_{++}$, where $K>0$.
	The monopolist's problem is to choose a price $a \in A \coloneqq \R_{++}$ to maximise profit
	%
	\begin{equation*}
		u(a,(c,-\eta)) \coloneqq (a-c) \cdot D(a,\eta) .
	\end{equation*}

@@

\begin{exercise}
	%
	\label{exercise:incr_covnex}
	%
	Let $A \subseteq \R$ and $\Theta \subseteq \R$ be non-empty and convex, let $u$ be a function $A \times \Theta \to \R$, and let $\phi$ be a function $\R \to \R$.

		\item Show that if $u(a,\theta) = \phi(a+\theta)$ for all $a \in A$ and $\theta \in \Theta$, then $u$ has increasing differences if and only if $\phi$ is convex.

		\item Show that if $u(a,\theta) = \phi(a-\theta)$ for all $a \in A$ and $\theta \in \Theta$, then $u$ has increasing differences if and only if $\phi$ is concave.

		\item Show that $(a,\theta) \mapsto -(a-\theta)^2$ has increasing differences. (This a common functional form: see e.g. \textcite{CrawfordSobel1982}.)

@@

	\item The action set $A$ can be multi-dimensional: $A \subseteq \R^L$ for $L \in \N$. In this case, an additional assumption is required, which says that the different dimensions of the action $a \in A$ are complementary with each other (not only with the parameter $\theta$). One such condition (which is stronger than necessary, but easy to interpret) is \emph{supermodularity} of $u(\cdot,\theta)$, which assuming twice continuous differentiability requires precisely that for each $\theta \in \Theta$,
	%
	\begin{multline*}
		u_{\ell m}(a_1,\dots,a_\ell,\dots,a_m,\dots,a_L,\theta) \geq 0
		\\
		\text{for all $\ell \neq m$ in $\{1,\dots,L\}$ and all $(a_1,\dots,a_\ell,\dots,a_m,\dots,a_L) \in A$.}
	\end{multline*}
	%
	My own experience is that such supermodularity-type assumptions rarely hold in economic applications, making this extension to multi-dimensional actions less useful than it sounds.

@@

	Recall the rudiments of dynamic programming \parencite[e.g.][]{StokeyLucasPrescott1989}. The Bellman equation is the following functional equation in an unknown function $f : \R_+ \to \R_+$:
	%
	\begin{equation*}
		f(w) =
		\max_{s \in [0,w]}
		\left[ U(w-s) + \delta \E\bigl( f\left( (1+R_1) s + Y_1 \right) \bigr) \right]
		\quad \text{for every $w \in \R_+$.}
	\end{equation*}
	%
	Under standard assumptions on $U$ and the distributions of $Y_1$ and $R_1$,%
		\footnote{For example, assume that $(Y_t,R_t)$ take values in a compact set and that $U$ is bounded and continuous, and appeal to the results in \textcite[][chapter~9]{StokeyLucasPrescott1989}.}
	the value function $V : \R_+ \to \R_+$ solves the Bellman equation, and is the only solution that is bounded; furthermore, optimal saving policies $\sigma : \R_+ \to \R_+$ are precisely those that satisfy
	%
	\begin{equation*}
		\sigma(w) \in \argmax_{s \in [0,w]}
		\left[ U(w-s) + \delta \E\bigl( V\left( (1+R_1) s + Y_1 \right) \bigr) \right] 
		\quad \text{for every $w \in \R_+$.}
	\end{equation*}
	%
	Prove the following.

@@

\begin{exercise}[based on \textcite{DekelQuahSinander}]
	%
	\label{exercise:adj}
	%
	Consider a model of costly adjustment. An agent chooses $a \in A \subseteq \R$, and her gross payoff is $u(a,\theta)$, where $\theta \in \Theta \subseteq \R^K$ is a parameter and $u$ has single-crossing differences. Initially, the parameter was $\theta_0 \in \Theta$, and the agent chose $a_0 \in \argmax_{a \in A} u(a,\theta_0)$. The parameter now increases to $\theta_1 \geq \theta_0$, and the agent adjusts her action in response. Adjustment is costly: in particular, adjusting by $\Delta = a_1 - a_0$ costs $c(\Delta) \geq 0$.  The agent chooses a new action to maximise her net payoff: $a_1 \in M_1 \coloneqq \argmax_{a \in A} [u(a,\theta_1)-c(a-a_0)]$. Assume that $M_1$ is non-empty.
